---
title: "Troubles with leading and trailing spaces"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

## 1.0 Setting up the environment

### 1.1 Loading packages

```{r}
library(tidyverse)
library(sf)
library(leaflet)
```

### 1.2 Setting the workspace
Linux has a little different structure than Windows

```{r}
setwd("~/dataStories/apartmentClimate/sourceData")
```

### 1.3 Importing the datasets
Apartment registration csv
Address points shapefile

```{r}
aptBldg <- read_csv("Apartment Building Registration Data.csv")
address <- st_read("ADDRESS_POINT_WGS84.shp")
```


## 2.0 Perform some data cleaning

* remove unneeded columns
* convert the address sf object to a dataframe for faster processing, as spatial operations will not be performed
* convert specific factor data to character
* create a address field that includes address number and street name, so that the two tables can be joined
* convert characters to upper case, which is needed for the join

```{r}
#Clear-up address data
#Remove unneeded columns
address <- address %>%
  select(ADDRESS, LONGITUDE, LATITUDE)

#Convert to data frame
addressDF <- as.data.frame(address)

#Convert factors to character
addressDF$ADDRESS <- as.character(addressDF$ADDRESS)
addressDF$LFNAME <- as.character(addressDF$LFNAME)
addressDF$LONUMSUF <- as.character(addressDF$LONUMSUF)
head(addressDF)

#Create combined address column
addressDF <- addressDF %>%
  unite(SITE_ADDRESS, c(ADDRESS, LFNAME), sep = " ")

#Convert charter to upper
addressDF <- mutate_all(addressDF, toupper)

#Remove white space from apartment dataset
aptBldg <- aptBldg %>%
  mutate(SITE_ADDRESS = str_squish(SITE_ADDRESS))
aptBldg

#Join datasets by site_address
aptBldg <- inner_join(aptBldg, addressDF)

#Convert long/lat back to double
aptBldg$LONGITUDE <- as.double(aptBldg$LONGITUDE)
aptBldg$LATITUDE <- as.double(aptBldg$LATITUDE)
```


```{r}
#Create sf obbject from long/lat coords
aptBldgGeo <- st_as_sf(aptBldg, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

aptBldgGeo %>%
  ggplot() +
  geom_sf()
```



```{r}
head(aptBldg)
```


```{r}
m <- leaflet(aptBldg) %>%
  addTiles() %>%
  addMarkers(lng = aptBldg$LONGITUDE, lat = aptBldg$LATITUDE,
             clusterOptions = markerClusterOptions(),
             popup = paste("Address:", aptBldg$SITE_ADDRESS, "<br>",
                           "Property owner:", aptBldg$PROP_MANAGEMENT_COMPANY_NAME, "<br>",
                           "Air conditioning:", aptBldg$AIR_CONDITIONING_TYPE, "<br>",
                           "Balconies:", aptBldg$BALCONIES, "<br>",
                           "Bike parking:", aptBldg$BIKE_PARKING, "<br>",
                           "Garbage chute:", aptBldg$GARBAGE_CHUTES, "<br>",
                           "Heating type", aptBldg$HEATING_TYPE, "<br>",
                           "Laundy room:", aptBldg$LAUNDRY_ROOM, "<br>",
                           "Lockers", aptBldg$LOCKER_OR_STORAGE_ROOM, "<br>",
                           "Num elevators:", aptBldg$NO_OF_ELEVATORS, "<br>",
                           "Num of storeys", aptBldg$NO_OF_STOREYS, "<br>",
                           "Num of units", aptBldg$NO_OF_UNITS, "<br>",
                           "Visitor parking:", aptBldg$VISITOR_PARKING, "<br>",
                           "Pets:", aptBldg$PETS_ALLOWED, "<br>",
                           "Year built:", aptBldg$YEAR_BUILT))
m
```


```{r}
#Some analytics
aptBldg %>%
  summarize(total_units = sum(NO_OF_UNITS))

aptBldg %>%
  group_by(BALCONIES) %>%
  count()

aptBldg %>%
  group_by(AIR_CONDITIONING_TYPE) %>%
  count()

aptBldg %>%
  group_by(HEATING_TYPE) %>%
  count()

aptBldg %>%
  group_by(YEAR_BUILT) %>%
  count() %>%
  arrange(desc(n))

aptBldg %>%
  filter(YEAR_BUILT > 1995) %>%
  group_by(YEAR_BUILT) %>%
  count() %>%
  arrange(desc(n))
```

```{r}
aptCount <- aptBldg %>%
  filter(YEAR_BUILT > 1900) %>%
  group_by(YEAR_BUILT) %>%
  count() %>%
  arrange(desc(n))
```

```{r}
aptCount
```


```{r}
aptCount %>%
  ggplot(aes(YEAR_BUILT, n)) +
  scale_y_continuous(minor_breaks = seq(0 , 400, 50), breaks = seq(0, 400, 50)) +
  scale_x_continuous(minor_breaks = seq(1900, 2020, 10), breaks = seq(1900, 2020, 10)) +
  geom_point()
```


```{r}
aptBldg %>%
  filter(YEAR_BUILT >= 1950 & YEAR_BUILT <= 1975) %>%
  count() %>%
  summarize(total_built = sum(n))
```
















